---
title: "NYPD Shooting"
output: html_document
date: "2025-06-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Import Libraries

```{r Load Packages, echo=TRUE, results='hide', message=FALSE}
if (!require(tidyverse)) install.packages("tidyverse")
library(tidyverse)

if (!require(janitor)) install.packages("janitor")
library(janitor)
```
### Read in and Clean Data
```{r import, echo=TRUE, results='hide', message=FALSE}

# Read in Data

url_in <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
nypd_shooting <- read_csv(url_in)

# Standardize Column Names

nypd_shooting <- clean_names(nypd_shooting)

# Convert dates

nypd_shooting <- nypd_shooting %>% 
  mutate(occur_date = mdy(occur_date))

# Further Cleaning for Later Analysis

nypd_shooting <- nypd_shooting %>%
  mutate(vic_age_group = recode(vic_age_group, "1022" = "10-22")) %>%
  filter(vic_age_group != "UNKNOWN")
```
### Vizualiton 1
``` {r Viz 1, echo=TRUE, results='hide', message=FALSE}
# Shootings by Borough

nypd_shooting %>%
  count(boro, sort = TRUE) %>%
  ggplot(aes(x = reorder(boro, n), y = n)) +
  geom_col(fill = "tomato") +
  coord_flip() +
  labs(title = "Shootings by Boro", x = "Boro", y = "Count") 

```

\newline
### Vizualization 2
``` {r Viz 2, echo=TRUE, results='hide', message=FALSE}
# Break Down of Shootings per Year by Borough

nypd_shooting %>%
  mutate(year = year(occur_date)) %>%
  count(year, boro) %>%
  ggplot(aes(x = factor(year), y = n, fill = boro)) +
  geom_col() +
  labs(
    title = "Annual Shootings in NYC by Borough",
    x = "Year", y = "Number of Shootings", fill = "Borough"
  ) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

\newline
### Vizualization 3
``` {r Viz 3, echo=TRUE, results='hide', message=FALSE}
# Victim Age Group Distribution by Borough

nypd_shooting %>%
  filter(vic_age_group != "10-22") %>% 
  count(boro, vic_age_group) %>%
  ggplot(aes(x = boro, y = n, fill = vic_age_group)) +
  geom_col() +
  labs(
    title = "Victim Age Group Distribution by Borough",
    x = "Borough", y = "Number of Victims", fill = "Victim Age Group"
  )

```

\newline
### Model
```{r model, warning=FALSE}

age_group_map <- c(
  "0-17" = 8, "18-24" = 21, "25-44" = 35, "45-64" = 55, "65+" = 70, 
  "UNKNOWN" = NA, "10-22" = 16
)

nypd_shooting <- nypd_shooting %>%
  mutate(
    year = year(occur_date),
    age_numeric = recode(vic_age_group, !!!age_group_map)
  )

avg_age_by_year <- nypd_shooting %>%
  filter(!is.na(age_numeric)) %>%
  group_by(year) %>%
  summarise(avg_age = mean(age_numeric), .groups = "drop")

mod_age <- lm(avg_age ~ year, data = avg_age_by_year)

avg_age_by_year %>%
  mutate(pred = predict(mod_age)) %>%
  ggplot(aes(x = year)) +
  geom_point(aes(y = avg_age), color = "blue") +
  geom_line(aes(y = pred), color = "red") +
  labs(
    title = "Average Victim Age Over Time",
    x = "Year", y = "Average Age"
  )

```

### Potential Bias

* Variable Selection Bias: The choice of which variables to include in the analysis can introduce bias by framing the story in a limited way. Key social, economic, or contextual factors may be left out—not because they are unimportant, but because they are not as readily available or measurable. This can lead to incomplete or skewed interpretations of complex issues like public health or law enforcement outcomes.

* Data Collection Bias: The dataset only includes incidents recorded by the NYPD. Unreported or misclassified shootings may not be captured, leading to underrepresentation of certain events or communities.

* Victim Demographic Uncertainty: Some demographic fields, such as victim age and sex, contain "UNKNOWN" or inconsistently entered values. Even after cleaning, this may skew analysis of victim characteristics.
